Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-11 paper.qmd
Updated .bib
Removed unnecessary references from .bib

At first glance, thinking of NLDR as a modeling technique might seem strange. It is a simplified representation or abstraction of a system, process, or phenomenon in the real world. The \pD{}   observations are the realization of the phenomenon, and the \kD{}   NLDR layout is the simplified representation. From a statistical perspective we can consider the distances between points in the \kD{}   layout to be variance that the model explains, and the (relative) difference with their distances in \pD{}   is the error, or unexplained variance. We can also imagine that the positioning of points in \gD{}    represent the fitted values, that will have some prescribed position in \pD{}   that can be compared with their observed values. This is the conceptual framework underlying the more formal versions of factor analysis [@cfa69] and MDS. (Note that, for this thinking the full \pD{}   data needs to be available, not just the interpoint distances.)

@@

Most NLDR methods only provide a \gD{} but if a \kD{} ($k>2$) layout is provided the approach developed here could be extended. Binning into cubes could be done in $3\text{-}D$ or higher, relatively easily, and used asa the basis for a wireframe of the fitted model. @barber1996 and the software @cxhull have algorithms for convex hulls, which \pD{} which serve as an inspiration. A simpler approach using $k$-means clustering to provide centroids could also be possible, but the complication would be to determine how to connect the centroids into an appropriate wireframe.

>>>>

At first glance, thinking of NLDR as a modeling technique might seem strange. It is a simplified representation or abstraction of a system, process, or phenomenon in the real world. The \pD{}   observations are the realization of the phenomenon, and the \kD{}   NLDR layout is the simplified representation. From a statistical perspective we can consider the distances between points in the \kD{}   layout to be variance that the model explains, and the (relative) difference with their distances in \pD{}   is the error, or unexplained variance. We can also imagine that the positioning of points in \gD{}    represent the fitted values, that will have some prescribed position in \pD{}   that can be compared with their observed values. This is the conceptual framework underlying the more formal versions of factor analysis [@joreskog1969] and MDS. (Note that, for this thinking the full \pD{}   data needs to be available, not just the interpoint distances.)

@@

Most NLDR methods only provide a \gD{} but if a \kD{} ($k>2$) layout is provided the approach developed here could be extended. Binning into cubes could be done in $3\text{-}D$ or higher, relatively easily, and used asa the basis for a wireframe of the fitted model. @barber1996 and the software @stephane2023 have algorithms for convex hulls, which \pD{} which serve as an inspiration. A simpler approach using $k$-means clustering to provide centroids could also be possible, but the complication would be to determine how to connect the centroids into an appropriate wireframe.
